Adaptive Video Streaming with Optimized Bitstream Extraction and PID-based Quality Control

### 2015-05-20 tmm.tex
add related work and refine the text

Bitstream extraction is the operation of choosing a subset of data packets from a whole SVC bitstream to form a sub-stream. Fig. \ref{fig:Bitstream-Extraction} illustrates the structure of an SVC stream and a possible way to extract a sub-stream from it. The data packets in the stream are divided into different layers. The temporal layer (T0$\sim$T3) reflects the reference or dependency relationship between the frames, shown by the arrows on the top pointing from the frame which is referenced to the frame which is dependent on the reference. The quality layer (Q0$\sim$Q2) indicates scalability of video quality within each frame. The dash line demonstrates the sample extraction which will keep the data packets below the line and discard all the packets above it. This is only one option of the many possible ways, because there are many combinations of packets to be discarded. Our goal, however, is to choose a way that maximums the video quality of the extracted sub-stream, under the constrain of a give bitrate.

@@

SVC bitstream extraction, as we can see, is substantially a combinatorial optimization problem. More specifically, it is in the form of ``0-1 Knapsack Problem" \cite{Knapsack}. 
Generally, the optimal solution of a typical ``0-1 Knapsack Problem" can be found using the dynamic programming algorithm. However, this is not operational for the bitstream extraction problem, mainly due to the dependence between packets. First, the value of a packet is not obvious and certain, since a packet's contribution to video quality may depend on the other packets. Second, the subset of packets to be extracted cannot be chosen arbitrarily, since if a packet is included, all the packets it depends on must also be included. In fact, to the best of our knowledge, no method presented in the literature claims to find the optimal solution of the SVC bitstream extraction problem. All the effort is made to obtain a solution that is as good as possible.

In \cite{Amonou07}, Amonou et al. proposed a bitstream extraction framework based on the concept of \textit{Quality Layers}. This framework has been adopted by the SVC reference software, i.e., JSVM  \cite{JSVM}, and achieves better R-D performance compared with the basic extractor of JSVM. In Amonou's framework, a Quality Layer value, which reflects a packet's R-D impact, is calculated and used for R-D optimized bitstream extraction. The process of obtaining distortion impact of every packet is extremely time consuming, and the distortion estimation accuracy can also be further improved. So other distortion/error models were proposed to either reduce the computational complexity or improve the estimation accuracy. Sun et al. \cite{Sun09} and Maani et al. \cite{Maani09} constructed models to calculate distortion based on the drift propagation between frames. Sun et al.'s method is almost as good as JSVM in regard of performance, but has much reduced complexity. Maani et al.'s model can achieve better estimation accuracy than JSVM; however, since it is a training-based model, more computation is needed to get robust model parameters.

@@

For a quality control algorithm, there are three targets that should be achieved. First is to ensure the continuity of video playing. In real-time streaming system, each data packet has its own playtime, which is also regarded as the deadline of the transmission. If the packet does not arrive at client side before the deadline, the media player will be paused due to lack of data, and that should be avoided. The second is to keep a smooth video quality. Although network conditions may change rapidly, it is necessary for the streamed video to maintain a relatively stable quality level, since frequent change of video quality will lead to poor user experience. The third target is to make the video quality as good as possible. In other words, the available bandwidth should be fully utilized. Suppose we keep sending video data in a low bitrate (with low video quality as well), neither the playback pause nor any quality fluctuation will likely happen. However, a low video quality is certainly not what we want.

@@

In DSS, quality control is implemented by monitoring packet delay (the difference between current time and the playing time of this packet). When a new packet is ready to be sent to streaming buffer, DSS checks the delay of it and compares this delay with some pre-set critical time thresholds to judge whether and how violently the quality change should be made. This algorithm can reduce quality fluctuation to some degree. However, it only provides conservative quality change, which makes its reaction so slow that sometimes media player may have already stopped due to lack of video data before quality drops. Furthermore, this algorithm has no knowledge of the bandwidth change so it cannot accurately adjust the video quality. Therefore, quality oscillation is inevitable in this algorithm.

The following of this paper will present our solutions for the above problems in detail. And we will combine our works together and lay a solid foundation for a practical adaptive video streaming system.

@@

Being able to adjust bitrate is the prerequisite for an adaptive video streaming system. For SVC videos, it means to extract a subset of packets from the whole bitstream. How to optimally choose this subset to achieve minimum distortion under the given bitrate is the key point of bitstream extraction problem. In this section, we propose a linear error mode and a greedy-like algorithm to solve this problem.

@@

The above model is called Linear Error Model (LEM). Note that in equations (\ref{eq:base0})-(\ref{eq:subset2}) the value of M and p remain the same, since in the case of quality scalability the enhancement packets are generated only by manipulating the transform coefficients, and not interfered with the motion compensation and static prediction.  This argument no longer holds for the other scalable dimensions, so the model does not simply apply for those cases. Further research and modification of this linear model for the spatial or temporal scalability would be our future work.

@@

@@

To implement our proposed adaptive streaming system requires the combination of the optimized bitstream extraction method and the PID-based quality control algorithm. The latter determines the current bitrate to follow the bandwidth change, and the former will ensure the sub-stream with the best possible quality is obtained and sent under that bitrate constraint. They work together to construct a system that delivers best quality of experience for the users. Choosing DSS as the implementation basis, we have integrated the proposed algorithms into it for the purpose of both research and application. In this section, we first present separate experiments to demonstrate the advantage of each proposed solution, and then give the result when they are combined together. Readers who are interested in the practical implementation can also visit the online video website http://www.7dlive.com for more information.

@@

In this paper, we propose an adaptive video streaming system with optimized bitstream extraction and PID-based quality control. The bitstream extraction method includes a linear error model to estimate the distortion caused by discarding any combination of packets and a greedy algorithm to assign priority to each data packet. The packets' priority values can then be stored in the bitstream and used for R-D optimized extraction. Experimental results show that the proposed method can achieve PSNR gain of up to 0.5dB compared to the JSVM QL extractor under the same bitrate constraint, also with the same computational complexity. The PID-based quality control algorithm has a flexible mechanism to determine the suitable quality level so that it gives an accurate and swift response to the bandwidth change. Comparing with the original algorithm in DSS, this algorithm has reduced quality fluctuation significantly to improve audience satisfaction. The SVC streaming system integrated with these two solutions performs well in automatically adapting to the user's bandwidth fluctuation, and is able to give the user a better watching experience.

Since the next generation video coding standard High Efficiency Video Coding (HEVC) \cite{HEVC} adopts the similar coding framework as its predecessor H.264/AVC, we expect to apply this paper's model and approach for bitstream extraction to the emerging HEVC standard and its scalable extension in our future work. As for the PID-based quality control algorithm, it is only network-concerned thus should be easily integrated into other streaming systems, e.g., simulcast system like DASH.

>>>>

Bitstream extraction is the operation of choosing a subset of data packets from a whole SVC bitstream to form a sub-stream. It makes the bitrate adjustable and is the prerequisite of an adaptive streaming system. Fig. \ref{fig:Bitstream-Extraction} illustrates the structure of an SVC stream and a possible way to extract a sub-stream from it. The data packets in the stream are divided into different layers. The temporal layer (T0$\sim$T3) reflects the reference or dependency relationship between the frames, shown by the arrows on the top pointing from the referenced frame to the frame which is dependent on it. The quality layer (Q0$\sim$Q2) indicates scalability of video quality within each frame, and packets in the higher quality layers (i.e., Q1 and Q2) are partially discarded to implement the extraction. For example, the dash line demonstrates the sample extraction which will keep the data packets below the line and discard all the packets above it. There are many combinations of packets to be discarded, which means there are many possible ways for an extraction. Our goal, however, is to choose a way that maximums the video quality of the extracted sub-stream, under the constraint of a given bitrate.

@@

As we can see, in its nature SVC bitstream extraction is a combinatorial optimization problem. More specifically, it is in the form of ``0-1 Knapsack Problem" \cite{Knapsack}. In the ``0-1 Knapsack Problem", given some objects, each with a weight and a value, one decides which object to include in the knapsack so that the total weight is no more than a given limit and the total value is as large as possible. Similarly, in bitstream extraction problem, each data packet can be treated as an item with certain weight (the packet's size) and value (its contribution to the video quality). The solving process is mainly to decide whether or not a packet should be included in the final extracted sub-stream, or in the term of ``0-1 Knapsack Problem", to choose which object to pack up.

For a typical ``0-1 Knapsack Problem", the optimal solution can generally be found using the dynamic programming algorithm. However, this is not operational for the bitstream extraction problem, mainly due to the dependence between the stream's data packets. First, the subset of packets to be extracted cannot be chosen arbitrarily, since if a packet is included, all the packets it depends on must also be included. For example, in Fig. \ref{fig:Bitstream-Extraction}, we cannot include the packet in Q2 of a frame without including the packet in Q1 of the same frame. Second, the value of a packet is not obvious and certain, since a packet's contribution to video quality may depend on the other packets. Unless we discard a packet and calculate the distortion it causes for real, there is no way to exactly tell its value. In fact, to the best of our knowledge, no method presented in the literature claims to find the optimal solution of the SVC bitstream extraction problem. All the effort is made to obtain a solution that is as good as possible.

In \cite{Amonou07}, Amonou et al. proposed a bitstream extraction framework based on the concept of \textit{Quality Layers}. This framework has been adopted by the SVC reference software, i.e., JSVM  \cite{JSVM}, and achieves better R-D performance compared with the basic extractor of JSVM. In Amonou's framework, a Quality Layer value, which reflects a packet's R-D impact, is calculated and used for R-D optimized bitstream extraction. The process of obtaining distortion impact of every packet is extremely time consuming, and the distortion estimation accuracy can also be further improved. So other distortion/error models were proposed to either reduce the computational complexity or improve the estimation accuracy. Sun et al. \cite{Sun09} and Maani et al. \cite{Maani09} constructed models to calculate distortion based on the drift propagation between frames. Sun et al.'s method is almost as good as JSVM in regard of performance, but has much reduced complexity. Maani et al.'s model can achieve better estimation accuracy than JSVM; however, since it is a training-based model, more computation is needed to get robust model parameters.
Ramanathan et al. \cite{Ramanathan12} proposed their concept and calculation method of the ``Quality Metadata'' to signal and use in bitstream extraction. Yang et al. \cite{Yang13} proposed a method based-on simulated annealing to solve the bitstream extraction problem. In both methods, the accuracy of distortion estimation is the key of the performance.

@@

Quality control is the second problem we will address in this paper. The result of a quality control algorithm determines the suitable bitrate for current transmission, which is directly related to the change of video quality perceived by the user. It should be noted that, the bitstream extraction is performed under constraint of that bitrate determined by quality control, and even in non-SVC streaming systems (DASH, for example), quality control is an important part that requires careful design.

For a quality control algorithm, there are three targets that should be achieved. First is to ensure the continuity of video playing. In real-time streaming system, each data packet has its own playtime, which is also regarded as the deadline of the transmission. If the packet does not arrive at client side before the deadline, the media player will be paused due to lack of data, and that should be avoided. The second is to keep a smooth video quality. Although network conditions may change rapidly, it is necessary for the streamed video to maintain a relatively stable quality level, since frequent change of video quality will lead to poor user experience. The third target is to make the video quality as good as possible. In other words, the available bandwidth should be fully utilized. Suppose we keep sending video data in a low bitrate (with low video quality as well), neither the playback will pause nor any quality fluctuation will happen. However, a low video quality is certainly not what we want.

@@

In the widely-used video streaming server DSS, quality control is implemented by monitoring packet delay (the difference between current time and the playing time of this packet). When a new packet is ready to be sent to streaming buffer, DSS checks the delay of it and compares this delay with some pre-set thresholds to judge whether and how violently the quality change should be made. This algorithm can reduce quality fluctuation to some degree. However, it only provides conservative quality change, which makes its reaction so slow that sometimes media player may have already stopped due to lack of video data before quality drops. Furthermore, this algorithm does not keep track of the bandwidth change so it cannot adjust the video quality accordingly. Quality oscillation might occur in this algorithm.

The following of this paper will present our solutions for the above problems in detail. The solutions then work together and serve as a solid foundation for a practical adaptive video streaming system.

@@

Being able to adjust bitrate is the prerequisite of an adaptive video streaming system. For SVC videos, it means to extract a subset of packets from the whole bitstream. How to optimally choose this subset to achieve minimum distortion under the given bitrate is the key point of bitstream extraction problem. In this section, we propose a linear error mode and a greedy-like algorithm to solve this problem.

@@

The above model is called Linear Error Model (LEM). Note that in equations (\ref{eq:base0})-(\ref{eq:subset2}) the value of M and p remain the same, since in the case of quality scalability the enhancement packets are generated only by manipulating the transform coefficients, and not interfered with the motion compensation and static prediction. This argument no longer holds for the other scalable dimensions (e.g., spatial scalability), so the model does not simply apply for those cases. Only quality scalable videos are used in our system, and further research for other scalable dimensions can be our future work.

@@

After each packet is assigned a priority value and optimized extraction is made possible, the video stream now can well adapt to different data rates. Next, we propose the quality control algorithm based on the PID control mechanism, which is responsible for determining a suitable data rate according to current bandwidth condition.

@@

To implement our proposed adaptive streaming system requires the combination of the optimized bitstream extraction method and the PID-based quality control algorithm. The latter determines the current bitrate to follow the bandwidth change, and the former will ensure the sub-stream with the best possible quality is obtained and sent under that bitrate constraint. They work together to deliver best quality of service for the user. Choosing DSS as the implementation basis, we have integrated the proposed algorithms into it for the purpose of both research and application. In this section, we first present separate experiments to show the advantage of each proposed solution, and then give the result when they are combined together. Readers who are interested in the practical implementation and demonstration can visit the online video website http://www.7dlive.com for more information.

@@

In this paper, we propose an adaptive video streaming system with optimized bitstream extraction and PID-based quality control. The bitstream extraction method includes a linear error model to estimate the distortion caused by discarding any combination of packets and a greedy algorithm to assign priority to each data packet. The priority values can then be stored in the bitstream and used for R-D optimized extraction. Experimental results show that the proposed method can achieve PSNR gain of up to 0.5dB compared to the JSVM QL extractor under the same bitrate constraint, also with the same computational complexity. The PID-based quality control algorithm has a flexible mechanism to determine the suitable quality level so that it gives an accurate and swift response to the bandwidth change. Comparing with the original algorithm in DSS, this algorithm has reduced quality fluctuation significantly to improve audience satisfaction. The SVC streaming system integrated with these two solutions performs well in automatically adapting to the user's bandwidth fluctuation, and is able to give the user a better watching experience.

Since the next generation video coding standard High Efficiency Video Coding (HEVC) \cite{HEVC} adopts the similar coding framework as its predecessor H.264/AVC, we expect to apply this paper's model and approach of bitstream extraction to the emerging HEVC standard and its scalable extension. As for the PID-based quality control algorithm, it is only network-concerned thus should be easily integrated into other streaming systems, e.g., simulcast system like DASH.
